---
title: "C04_assignment"
output: html_document
---

# Setup 

As always, we start by running our setup function. Start RStudio/R, and reload your project with the menu `File > Recent Projects`. 

```{r setup}
source("setup.R")
```

# Load data - choose a month and sampling approach

Let's load what we need to build a model" the configuration and the complete covariate data.  Notice that we (a) transform month to a simple numeric value (it's easier to work with as a number) and (b) select the covariates that are included in our configuration plus "class".

You may also catch that we log scale `depth` and `Xbtm`.  This reduces the numeric range of each to be more comparable to those of the other covariates. This is a common trick when dealing with data that ranges over orders of magnitude compared to companion data.


```{r load_data}
cfg = read_configuration(scientificname = "Mustelus canis", version = "v1")
model_input = read_model_input(scientificname = "Mustelus canis", 
                               version = "v1",
                               log_me = c("depth", "Xbtm")) |>
  dplyr::mutate(month = month_as_number(.data$month)) |>
  select(all_of(c("class", cfg$keep)))
```

# Splitting the data into groups

Here we want to split the data into two groups: data for training the model during creation and data for testing the model after we have created it. We'll aim for 80% of the data to train on and reserving 20% of the data for later testing. 

After that, we want to split the **training** data so that we have groups, or "folds", across the spatial domain. This will look approximately like a checkerboard where we (potentially) have a balance of observations and background represented, but it is possible to have strong imbalance in any of the checkerboard squares. That's OK.

## Initial split into training and testing groups

```{r initial_split}
model_input_split = spatial_initial_split(model_input, 
                        prop = 1 / 5,     # 20% for testing
                        strategy = spatial_block_cv) # see ?spatial_block_cv
model_input_split
```

We can see that there are 5 times more points (observations and background) reserved for training.  Now we can look at the spatial distribution of the two groups.

```{r initial_split_plot}
autoplot(model_input_split)
```

## Split the training group into regional folds

Next we take just the **training** data and split that into a set cross-validation folds.  We do this to improve our model by making multiple spatially distributed training mini-datasets.

```{r cv_training}
tr_data = training(model_input_split)
cv_tr_data <- spatial_block_cv(tr_data,
  v = 5,     
  cellsize = grid_cellsize(model_input),
  offset = grid_offset(model_input) + 0.00001
)
autoplot(cv_tr_data)
```

```{r recipe}
one_row_of_training_data = dplyr::slice(tr_data,1)
rec = recipe(one_row_of_training_data, formula = class ~ .)
rec
```
This print out provides a very high level summary - all of the details are glossed over. To get a more detailed summary use the `summary()` function.  Note that the coordinates (longitude and latitude) are detected and labeled as `coords` and not predictors. There are ways around that if you need to use longitude or latitude as predictor variables.

```{r recipe_summary}
summary(rec)
```

```{r make_workflow}
wflow = workflow_set(
  
  preproc = list(default = rec), # not much happening in our preprocessor
  
  models = list(                 # but we have 4 models to add
    
      # very simple - nothing to tune
      glm = logistic_reg(
          mode = "classification") |>
        set_engine("glm"),
      
      # two knobs to tune
      rf = rand_forest(
          mtry = tune(),
          trees = tune(),
          mode = "classification") |>
        set_engine("ranger", 
                   importance = "impurity"),
      
      # so many things to tune!
      btree = boost_tree(
          mtry = tune(), 
          trees = tune(), 
          tree_depth = tune(), 
          learn_rate = tune(), 
          loss_reduction = tune(), 
          stop_iter = tune(),
          mode = "classification") |>
        set_engine("xgboost"),
    
      # just two again
      maxent = maxent(
          feature_classes = tune(),
          regularization_multiplier = tune(),
          mode = "classification") |>
        set_engine("maxnet")
  )
)
wflow
```


```{r metrics}
metrics = sdm_metric_set(yardstick::accuracy)
metrics
```

```{r fit, warning = FALSE}
wflow <- wflow |>
  workflow_map("tune_grid",
    resamples = cv_tr_data, 
    grid = 3,
    metrics = metrics, 
    verbose = TRUE)
```
Here's a quick plot of the outputs, with each metric being showcased.
```{r plot_wflow}
autoplot(wflow)
```

```{r select_best}
model_fits = workflowset_selectomatic(wflow, model_input_split,
                                  filename = "Mustelus_canis-v1-model_fits",
                                  path = data_path("models"))
model_fits
```

```{r model_fit_metrics}
model_fit_metrics(model_fits)
```

```{r model_fit_confmat}
model_fit_confmat(model_fits)
```

```{r model_fit_roc_auc}
model_fit_roc_auc(model_fits)
```

```{r model_fit_vip}
model_fit_varimp_plot(model_fits)
```

```{r random_forest}
rf = model_fits |>
  filter(wflow_id == "default_rf")
rf
```

```{r rf_splits}
autoplot(rf$splits[[1]])
```

```{r rf_metrics}
rf$.metrics[[1]]
```

```{r rf_preds}
rf$.predictions[[1]]
```

```{r rf_workflow}
rf$.workflow[[1]]
```

```{r pd_plot}
model_fit_pdp(model_fits, wid = "default_btree", title = "Boosted Tree")
```