---
title: "My SecondSpecies- Smooth Dogfish (Mustelus canis) "
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#begin with setup 
```{r}
source("setup.R")
```


#load and filter data for new species 
```{r}
squalus = fetch_obis(scientificname = "Mustelus canis") |>
  print()
summary(squalus)
count(squalus, basisOfRecord)
count(squalus, month)
squalus = squalus |>
  filter(!is.na(eventDate))
summary(squalus)
human = obs |>
  filter(basisOfRecord == "HumanObservation") |>
  slice(1) |>
  browse_obis()
preserved = obs |>
  filter(basisOfRecord == "PreservedSpecimen") |>
  slice(1) |>
  browse_obis()
checklist = obs |>
  filter(basisOfRecord == "NomenclaturalChecklist") |>
  slice(1) |>
  browse_obis()
occurrence = obs |>
  filter(basisOfRecord == "Occurrence") |>
  slice(1) |>
  browse_obis()
summary(obs)
obs = obs |>
  filter(!is.na(eventDate))
summary(obs)
obs |>
  filter(is.na(individualCount)) |>
  slice(1) |>
  browse_obis()
obs = obs |>
  filter(!is.na(individualCount))
summary(obs)
ggplot(data = obs,
       mapping = aes(x = year)) + 
  geom_bar() + 
  geom_vline(xintercept = c(1982, 2013), linetype = "dashed") + 
  labs(title = "Counts per year")
obs = obs |>
  filter(year >= 1970)
dim(obs)
ggplot(data = obs,
       mapping = aes(x = month)) + 
  geom_bar() + 
  labs(title = "Counts per month")
obs = obs |>
  mutate(month = factor(month, levels = month.abb))

ggplot(data = obs,
       mapping = aes(x = month)) + 
  geom_bar() + 
  labs(title = "Counts per month")
db = brickman_database() |>
  filter(scenario == "STATIC", var == "mask")
mask = read_brickman(db)
mask
plot(mask, breaks = "equal", axes = TRUE, reset = FALSE)
plot(st_geometry(obs), pch = ".", add = TRUE)
hitOrMiss = extract_brickman(mask, obs)
hitOrMiss
count(hitOrMiss, value)
dim_start = dim(obs)
obs = obs |>
  filter(!is.na(hitOrMiss$value))
dim_end = dim(obs)

dropped_records = dim_start[1] - dim_end[1]
dropped_records
```

#load the data from the observations function
```{r}
coast = read_coastline()
obs = read_observations(scientificname = "Mustelus canis")
db = brickman_database() |>
  filter(scenario == "STATIC", var == "mask")
mask = read_brickman(db)

mask = read_brickman(db)
db_covariates = brickman_database() |>
  filter(scenario == "PRESENT",
         interval == "mon",
         var %in% c("SST", "U", "SSS")
         )
covariates <- read_brickman(db_covariates)
```

#add labels and notation

```{r}
LON0 = -67
LAT0 = 46
all_counts = count(st_drop_geometry(obs), month) # counting is faster without spatial baggage
all_counts
```

#limit the amount of observations so that there is only one per month

```{r}
thinned_obs = sapply(month.abb,
               function(mon){ 
                 thin_by_cell(obs |> filter(month == mon), mask)
               }, simplify = FALSE) |>
  dplyr::bind_rows() 

# another count
thinned_counts = count(st_drop_geometry(thinned_obs), month)

ggplot() +
  geom_sf(data = thinned_obs, 
          alpha = 0.2, 
          shape = "circle small", 
          size = 1) +
  geom_sf(data = coast, col = "orange") +
  geom_text(data = thinned_counts,
            mapping = aes(x = LON0, 
                          y = LAT0, 
                          label = sprintf("n: %i", .data$n)),
                          size = 3) + 
  labs(x = "Longitude", y = "Latitude", title = "Thinned observations") +
  facet_wrap(~month)
```

#create a bias map using all data to understand the quantity of total observations in each cell before the data was thinned

```{r}
bias_map = rasterize_point_density(obs, mask) # <-- note the original observations

ggplot() +
  geom_stars(data = bias_map, aes(fill = count)) +
  scale_fill_viridis_b(na.value = "transparent") +
  geom_sf(data = coast, col = "orange") + 
  labs(x = "Longitude", y = "Latitude", title = "Bias map using all observations")
```

#average data per month

```{r}
nback_avg = mean(all_counts$n) |>
  round()
nback_avg
```

#randomly select background points using thinned observations and bias map

```{r}
obsbkg = sapply(month.abb,
    function(mon){ 
      sample_background(thinned_obs |> filter(month == mon), # <- just this month
                       bias_map,
                       method = "bias",  # <-- it needs to know it's a bias map
                       return_pres = TRUE, # <-- give me the obs back, too
                       n = nback_avg) |>   # <-- how many points
        mutate(month = mon, .before = 1)
    }, simplify = FALSE) |>
  bind_rows() |>
  mutate(month = factor(month, levels = month.abb))
obsbkg 
```

#confirm that the number of points used is the same as the calculated average
```{r}
count(st_drop_geometry(obsbkg), month, class)
```

#plot the data to understand where the background and presence data points are in realtion to eachother

```{r}
ggplot() +
  geom_sf(data = obsbkg, 
          mapping = aes(col = class),
          alpha =  0.4, shape = "circle small", size = 1) +
  geom_sf(data = coast, col = "orange")  + 
  labs(x = "Longitude", y = "Latitude", title = "All") +   
  theme_bw() +  # <- make a simple white background
  scale_fill_okabe_ito() +  # <-- colorblind friendly for N Record
  facet_wrap(~month)
```

#save the function so that it can be used in the future

```{r}
write_model_input(obsbkg, scientificname = "Mustelus canis")
```

#this function reads model input files

```{r}
x = read_model_input(scientificname = "Mustelus canis")
```

#create table with new data points and multiple variables

```{r}
result = x |>
  
  group_by(month) |>
  group_map(function(rows, key) {
  
    rows_sf = sf::st_as_sf(rows, coords = c("lon", "lat"), crs = 4326)

    cv_metadata = db_covariates |> filter(var %in% c("SST", "SSS", "U"))
    cv_spatial = read_brickman(cv_metadata) 
    
    vals = extract_brickman(cv_spatial, rows_sf, form = "wide") |> 
      select(-.id)
    
    return(vals)
  }, .keep = TRUE) |> 
  bind_rows()
  
print(result)
```